---
format:
  html:
    code-overflow: wrap
engine: knitr
execute:
  eval: true
  error: true
title: "Polars vs Tidyverse"
description: "polars side by side with the tidyverse"
---

<style>
pre, pre code, pre.sourceCode { font-size: 0.82em; line-height: 1.35; }
div.compare { display: flex; gap: 1.5rem; }
div.compare-col { flex: 1 1 0; min-width: 0; }
/* R text output (cell-output-stdout) gets a cream background by default
   from bootstrap's .cell-output.cell-output-stdout rule; override it so
   the tidyverse column matches the Polars (transparent) output. */
.cell-output.cell-output-stdout,
.cell-output.cell-output-stdout pre { background-color: transparent !important; }
@media screen {
  pre > code.sourceCode {
    white-space: pre-wrap;
    overflow-wrap: anywhere;
    word-break: break-word;
  }
  div.sourceCode { overflow-x: hidden; }
}
</style>

## Create the data

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
import polars as pl

adsl = pl.DataFrame({
    "USUBJID": ["U01", "U02", "U03", "U04"],
    "ARM":     ["DRUG A", "DRUG B", "DRUG A", "DRUG B"],
    "AGE":     [55, 62, 48, 70],
    "SEX":     ["M", "F", "M", "F"],
})

advs = pl.DataFrame({
    "USUBJID": ["U01","U01","U02","U02","U03","U03","U04","U04"],
    "ARM":     ["DRUG A","DRUG A","DRUG B","DRUG B","DRUG A","DRUG A","DRUG B","DRUG B"],
    "AGE":     [55, 55, 62, 62, 48, 48, 70, 70],
    "SEX":     ["M","M","F","F","M","M","F","F"],
    "PARAM":   ["Systolic BP","Diastolic BP","Systolic BP","Diastolic BP","Systolic BP","Diastolic BP","Systolic BP","Diastolic BP"],
    "PARAMCD": ["SYSBP","DIABP","SYSBP","DIABP","SYSBP","DIABP","SYSBP","DIABP"],
    "AVAL":    [120, 80, 130, 85, 110, 70, 140, 90],
    "AVISIT":  ["Visit 1","Visit 1","Visit 1","Visit 1","Visit 2","Visit 2","Visit 2","Visit 2"],
    "AVISITN": [1, 1, 1, 1, 2, 2, 2, 2],
    "DTC":     ["2024-01-15","2024-01-15","2024-02-20","2024-02-20","2024-03-10","2024-03-10","2024-04-05","2024-04-05"],
})
adsl
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
#| message: false
#| warning: false
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
library(readr)

adsl <- tibble::tibble(
  USUBJID = c("U01", "U02", "U03", "U04"),
  ARM     = c("DRUG A", "DRUG B", "DRUG A", "DRUG B"),
  AGE     = c(55, 62, 48, 70),
  SEX     = c("M", "F", "M", "F")
)

advs <- tibble::tibble(
  USUBJID = c("U01","U01","U02","U02","U03","U03","U04","U04"),
  ARM     = c("DRUG A","DRUG A","DRUG B","DRUG B","DRUG A","DRUG A","DRUG B","DRUG B"),
  AGE     = c(55, 55, 62, 62, 48, 48, 70, 70),
  SEX     = c("M","M","F","F","M","M","F","F"),
  PARAM   = c("Systolic BP","Diastolic BP","Systolic BP","Diastolic BP","Systolic BP","Diastolic BP","Systolic BP","Diastolic BP"),
  PARAMCD = c("SYSBP","DIABP","SYSBP","DIABP","SYSBP","DIABP","SYSBP","DIABP"),
  AVAL    = c(120, 80, 130, 85, 110, 70, 140, 90),
  AVISIT  = c("Visit 1","Visit 1","Visit 1","Visit 1","Visit 2","Visit 2","Visit 2","Visit 2"),
  AVISITN = c(1, 1, 1, 1, 2, 2, 2, 2),
  DTC     = c("2024-01-15","2024-01-15","2024-02-20","2024-02-20","2024-03-10","2024-03-10","2024-04-05","2024-04-05")
)
adsl
```
:::
:::

## Verbs

### Select columns

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
adsl.select("USUBJID", "ARM", "AGE")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
adsl |> select(USUBJID, ARM, AGE)
```
:::
:::

### Drop columns

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
adsl.drop("SEX")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
adsl |> select(-SEX)
```
:::
:::

### Drop by rule, not by name

R's `where(is.numeric)` and `starts_with()` have direct counterparts: `cs.numeric()`
and `cs.starts_with()`, composed with `|` and `&`.

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
import polars.selectors as cs

adsl.select(cs.numeric() | cs.starts_with("U"))
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
adsl |> select(where(is.numeric) | starts_with("U"))
```
:::
:::

### Rename columns

The dictionary is keyed old-to-new, the opposite of `rename()`'s `new = old`.

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
adsl.rename({"AGE": "AGE_N"})
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
adsl |> rename(AGE_N = AGE)
```
:::
:::

### Filter rows

Conditions are expressions joined with `&`, `|`, `~`; `is_in` is `%in%`. Note that
polars has no unquoted column names - every reference goes through `pl.col`.

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.filter(
    (pl.col("ARM") == "DRUG A") & (pl.col("AVISIT").is_in(["Visit 1", "Visit 2"]))
)
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |> filter(ARM == "DRUG A" & AVISIT %in% c("Visit 1", "Visit 2"))
```
:::
:::

### Add or replace columns

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
(
    advs
    .with_columns(doubled=pl.col("AVAL") * 2)
    .with_columns(squared=pl.col("doubled") ** 2)
    .select("USUBJID", "AVISIT", "AVAL", "doubled", "squared")
)
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |>
  mutate(
    doubled = AVAL * 2,
    squared = doubled^2
  ) |>
  select(USUBJID, AVISIT, AVAL, doubled, squared)
```
:::
:::

### Conditional values (case_when)

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.with_columns(
    response=pl.when(pl.col("AVAL") > 120)
    .then(pl.lit("HIGH"))
    .otherwise(pl.lit("NORMAL"))
).select("USUBJID", "AVAL", "response")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |>
  mutate(response = case_when(AVAL > 120 ~ "HIGH", TRUE ~ "NORMAL")) |>
  select(USUBJID, AVAL, response)
```
:::
:::

### Sort rows

The default placement of missing values differs: `arrange()` sorts `NA` last,
`sort()` puts nulls first unless `nulls_last=True`.

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
adsl.sort("AGE", descending=True, nulls_last=True)
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
adsl |> arrange(desc(AGE))
```
:::
:::

### Slice by row position

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
adsl.head(3)
adsl.tail(2)
adsl.slice(1, 2)
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
adsl |> slice_head(n = 3)
adsl |> slice_tail(n = 2)
adsl |> slice(2:3)
```
:::
:::

### Distinct

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.select("USUBJID", "ARM").unique().sort("USUBJID")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |> distinct(USUBJID, ARM) |> arrange(USUBJID)
```
:::
:::

### Type conversion

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
adsl.with_columns(id=pl.col("USUBJID").str.extract(r"(\d+)").cast(pl.Int32)).select("USUBJID", "id")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
adsl |> mutate(id = as.integer(str_extract(USUBJID, "\\d+"))) |> select(USUBJID, id)
```
:::
:::

### Batch rename (prefix / suffix)

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
adsl.select(pl.col("AGE").name.prefix("BASE_"))
adsl.select(pl.all().name.suffix("_raw"))
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
adsl |> select(AGE) |> rename_with(~ str_c("BASE_", .x))
adsl |> rename_with(~ str_c(.x, "_raw"))
```
:::
:::

## Missing values

### Check for missing values

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.null_count()
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |> summarise(across(everything(), ~ sum(is.na(.))))
```
:::
:::

### Drop rows with a missing value

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
miss = advs.with_columns(AVAL=pl.when(pl.col("USUBJID") == "U02").then(None).otherwise(pl.col("AVAL")))
miss.drop_nulls("AVAL").select("USUBJID", "AVAL")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
miss <- advs |> mutate(AVAL = ifelse(USUBJID == "U02", NA, AVAL))
miss |> drop_na(AVAL) |> select(USUBJID, AVAL)
```
:::
:::

### Fill with a constant

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
miss.with_columns(filled=pl.col("AVAL").fill_null(0)).select("USUBJID", "AVAL", "filled")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
miss |>
  mutate(filled = replace_na(AVAL, 0)) |>
  select(USUBJID, AVAL, filled)
```
:::
:::

### Carry the last value forward

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
miss.with_columns(
    filled=pl.col("AVAL").fill_null(strategy="forward")
).select("USUBJID", "AVISIT", "AVAL", "filled")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
miss |>
  fill(AVAL, .direction = "down") |>
  select(USUBJID, AVISIT, AVAL)
```
:::
:::

### Fill within a group

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
miss.with_columns(
    filled=pl.col("AVAL").fill_null(strategy="forward").over("USUBJID")
).select("USUBJID", "AVISIT", "AVAL", "filled")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
miss |>
  group_by(USUBJID) |>
  fill(AVAL, .direction = "down") |>
  ungroup() |>
  select(USUBJID, AVISIT, AVAL)
```
:::
:::

### Coalesce

`pl.coalesce` takes columns, `coalesce()` takes vectors.

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
co = pl.DataFrame({"a": [1, None, 3], "b": [10, 20, None]})
co.with_columns(c=pl.coalesce("a", "b"))
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
co <- tibble(a = c(1, NA, 3), b = c(10, 20, NA))
co |> mutate(c = coalesce(a, b))
```
:::
:::

### Delete empty rows

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
blank = pl.DataFrame({"USUBJID": [None], "AVAL": [None]})
pl.concat([miss.select("USUBJID", "AVAL"), blank], how="vertical").drop_nulls()
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
blank <- tibble(USUBJID = NA_character_, AVAL = NA_real_)
bind_rows(miss |> select(USUBJID, AVAL), blank) |> drop_na()
```
:::
:::

## Group and window

### Summarise by group


::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.group_by("ARM").agg(
    n=pl.len(),
    mean_age=pl.col("AGE").mean(),
    mean_aval=pl.col("AVAL").mean(),
)
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |>
  group_by(ARM) |>
  summarise(
    n = n(),
    mean_age = mean(AGE),
    mean_aval = mean(AVAL)
  )
```
:::
:::

### Common statistics

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.select(
    mean=pl.col("AVAL").mean(),
    median=pl.col("AVAL").median(),
    sd=pl.col("AVAL").std(),
    p75=pl.col("AVAL").quantile(0.75),
    total=pl.col("AVAL").sum(),
)
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |>
  summarise(
    mean = mean(AVAL),
    median = median(AVAL),
    sd = sd(AVAL),
    p75 = quantile(AVAL, 0.75),
    total = sum(AVAL)
  )
```
:::
:::

### Summarise many columns at once (across)

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.group_by("ARM").agg(
    pl.col("AGE", "AVAL").mean().name.prefix("mean_"),
    pl.col("AGE", "AVAL").std().name.prefix("sd_"),
)
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |>
  group_by(ARM) |>
  summarise(across(c(AGE, AVAL), list(mean = mean, sd = sd)))
```
:::
:::

### Count rows per group

`len()` is `n()`.

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.group_by("ARM").len()
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |> count(ARM)
```
:::
:::

### Window value (recycle an aggregate)

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.with_columns(
    subj_mean=pl.col("AVAL").mean().over("USUBJID")
).select("USUBJID", "AVISIT", "AVAL", "subj_mean")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |>
  group_by(USUBJID) |>
  mutate(subj_mean = mean(AVAL)) |>
  ungroup() |>
  select(USUBJID, AVISIT, AVAL, subj_mean)
```
:::
:::

### Running total within a group

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.with_columns(
    cum_aval=pl.col("AVAL").cum_sum().over("USUBJID")
).select("USUBJID", "AVISIT", "AVAL", "cum_aval")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |>
  group_by(USUBJID) |>
  mutate(cum_aval = cumsum(AVAL)) |>
  ungroup() |>
  select(USUBJID, AVISIT, AVAL, cum_aval)
```
:::
:::

### Lag and lead

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.with_columns(
    prev=pl.col("AVAL").shift(1).over("USUBJID"),
    lead_val=pl.col("AVAL").shift(-1).over("USUBJID"),
).select("USUBJID", "AVISIT", "AVAL", "prev", "lead_val")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |>
  group_by(USUBJID) |>
  mutate(
    prev = lag(AVAL),
    lead_val = lead(AVAL)
  ) |>
  ungroup() |>
  select(USUBJID, AVISIT, AVAL, prev, lead_val)
```
:::
:::

### First and last row of each group

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
(
    advs
    .with_columns(
        rn=pl.col("AVAL").rank(method="ordinal").over("USUBJID", "AVISITN"),
        is_first=pl.col("AVAL").rank(method="ordinal").over("USUBJID", "AVISITN") == 1,
        is_last=pl.col("AVAL").rank(method="ordinal").over("USUBJID", "AVISITN") == pl.len().over("USUBJID", "AVISITN"),
    )
    .select("USUBJID", "AVISIT", "rn", "is_first", "is_last")
)
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |>
  group_by(USUBJID, AVISITN) |>
  mutate(
    rn = row_number(),
    is_first = row_number() == 1,
    is_last = row_number() == n()
  ) |>
  ungroup() |>
  select(USUBJID, AVISIT, rn, is_first, is_last)
```
:::
:::

### Row-wise sum of several columns

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.with_columns(total=pl.sum_horizontal("AGE", "AVAL")).select("USUBJID", "AGE", "AVAL", "total")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |>
  mutate(total = AGE + AVAL) |>
  select(USUBJID, AGE, AVAL, total)
```
:::
:::

## Join and reshape

### Left join

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
arms = pl.DataFrame({"ARM": ["DRUG A", "DRUG B"], "ARMCD": ["A", "B"]})
adsl.join(arms, on="ARM", how="left")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
arms <- tibble(ARM = c("DRUG A", "DRUG B"), ARMCD = c("A", "B"))
adsl |> left_join(arms, by = "ARM")
```
:::
:::

### Set operations

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
a = pl.DataFrame({"USUBJID": ["U01", "U02", "U03"]})
b = pl.DataFrame({"USUBJID": ["U02", "U03", "U04"]})
a.filter(pl.col("USUBJID").is_in(b["USUBJID"].to_list()))
a.join(b, on="USUBJID", how="anti")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
a <- tibble(USUBJID = c("U01", "U02", "U03"))
b <- tibble(USUBJID = c("U02", "U03", "U04"))
a |> filter(USUBJID %in% b$USUBJID)
a |> anti_join(b, by = "USUBJID")
```
:::
:::

### Pivot long to wide

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
wide = advs.pivot(index="USUBJID", on="PARAMCD", values="AVAL")
wide
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
wide <- advs |> pivot_wider(names_from = PARAMCD, values_from = AVAL)
wide
```
:::
:::

### Pivot wide to long

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
wide.unpivot(index="USUBJID", on=["SYSBP","DIABP"], variable_name="PARAMCD", value_name="AVAL")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
wide |> pivot_longer(c(SYSBP, DIABP), names_to = "PARAMCD", values_to = "AVAL")
```
:::
:::

### Stack frames vertically

`concat([...], how="vertical")` is `bind_rows`.

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
top = advs.head(2)
bot = advs.tail(2)
pl.concat([top, bot], how="vertical")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
top <- advs |> head(2)
bot <- advs |> tail(2)
bind_rows(top, bot)
```
:::
:::

### Stack frames side by side

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
left = adsl.select("USUBJID", "ARM")
right = adsl.select("AGE", "SEX")
pl.concat([left, right], how="horizontal_extend")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
left <- adsl |> select(USUBJID, ARM)
right <- adsl |> select(AGE, SEX)
bind_cols(left, right)
```
:::
:::

## Strings

### Change case

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.select(
    upper=pl.col("ARM").str.to_uppercase(),
    lower=pl.col("PARAMCD").str.to_lowercase(),
    title=pl.col("PARAM").str.to_titlecase(),
)
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |> transmute(
  upper = str_to_upper(ARM),
  lower = str_to_lower(PARAMCD),
  title = str_to_title(PARAM)
)
```
:::
:::

### Trim and measure length

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.select(
    n=pl.col("PARAM").str.len_chars(),
    trimmed=pl.col("PARAM").str.strip_chars(),
)
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |> transmute(
  n = str_length(PARAM),
  trimmed = str_trim(PARAM)
)
```
:::
:::

### Test for a pattern

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.filter(pl.col("PARAMCD").str.contains(r"^S"))
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |> filter(str_detect(PARAMCD, "^S"))
```
:::
:::

### Extract the first match

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.with_columns(num=pl.col("USUBJID").str.extract(r"(\d+)")).select("USUBJID", "num")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |> mutate(num = str_extract(USUBJID, "\\d+")) |> select(USUBJID, num)
```
:::
:::

### Replace a pattern

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.with_columns(arm=pl.col("ARM").str.replace(r"DRUG ", "")).select("ARM", "arm")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |> mutate(arm = str_remove(ARM, "DRUG ")) |> select(ARM, arm)
```
:::
:::

### Split and merge

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.select(
    parts=pl.col("PARAM").str.split(" "),
    merged=pl.col("USUBJID") + pl.lit("-") + pl.col("AVISIT"),
)
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |> transmute(
  parts = str_split(PARAM, " "),
  merged = str_c(USUBJID, "-", AVISIT)
)
```
:::
:::

### Slice and pad

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.select(
    sliced=pl.col("PARAMCD").str.slice(0, 3),
    padded=pl.col("USUBJID").str.zfill(6),
)
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |> transmute(
  sliced = str_sub(PARAMCD, 1, 3),
  padded = str_pad(USUBJID, width = 6, side = "left", pad = "0")
)
```
:::
:::

## Dates

### Parse a date, datetime or time

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.with_columns(
    date=pl.col("DTC").str.to_date(),
).select("USUBJID", "date")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |> mutate(date = ymd(DTC)) |> select(USUBJID, date)
```
:::
:::

### Pull out parts

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.with_columns(
    d=pl.col("DTC").str.to_date(),
).select(
    year=pl.col("d").dt.year(),
    month=pl.col("d").dt.month(),
    day=pl.col("d").dt.day(),
    wday=pl.col("d").dt.weekday(),
)
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |>
  mutate(d = ymd(DTC)) |>
  transmute(
    year = year(d),
    month = month(d),
    day = mday(d),
    wday = wday(d)
  )
```
:::
:::

### Format a date as text

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.with_columns(
    txt=pl.col("DTC").str.to_date().dt.strftime("%d %b %Y")
).select("DTC", "txt")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |> mutate(txt = format(ymd(DTC), "%d %b %Y")) |> select(DTC, txt)
```
:::
:::

### Truncate to a period

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
advs.with_columns(
    month_start=pl.col("DTC").str.to_date().dt.truncate("1mo"),
).select("DTC", "month_start")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |> mutate(month_start = floor_date(ymd(DTC), "month")) |> select(DTC, month_start)
```
:::
:::

### Month start and end

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
(
    advs
    .with_columns(d=pl.col("DTC").str.to_date())
    .with_columns(
        ms=pl.col("d").dt.month_start(),
        me=pl.col("d").dt.month_end(),
    )
    .select("DTC", "ms", "me")
)
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
advs |>
  mutate(
    d = ymd(DTC),
    ms = floor_date(d, "month"),
    me = ceiling_date(d, "month") - days(1)
  ) |>
  select(DTC, ms, me)
```
:::
:::

### Add or subtract a period

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
jan = pl.DataFrame({"d": ["2024-01-31"]}).with_columns(d=pl.col("d").str.to_date())
jan.with_columns(
    next_month=pl.col("d").dt.offset_by("1mo"),
    prev_month=pl.col("d").dt.offset_by("-1mo"),
)
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
jan <- tibble(d = ymd("2024-01-31"))
jan |>
  mutate(
    next_month = d %m+% months(1),
    prev_month = d %m-% months(1)
  )
```
:::
:::

### Difference between dates

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
jan = pl.DataFrame({"d1": ["2024-01-01"], "d2": ["2024-04-01"]}).with_columns(
    pl.col("d1", "d2").str.to_date()
)
jan.with_columns(days=pl.col("d2") - pl.col("d1"))
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
jan <- tibble(d1 = ymd("2024-01-01"), d2 = ymd("2024-04-01"))
jan |> mutate(days = d2 - d1)
```
:::
:::

### Generate a sequence of dates

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
pl.DataFrame({"d": pl.date_range(pl.date(2024, 1, 1), pl.date(2024, 1, 5), "1d", eager=True)})
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
tibble(d = seq(ymd("2024-01-01"), ymd("2024-01-05"), by = "day"))
```
:::
:::

## Input / output

### Read and write CSV

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
#| eval: false
# write
advs.write_csv("advs.csv")
# read
advs = pl.read_csv("advs.csv", try_parse_dates=True)
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
#| eval: false
# write
advs |> write_csv("advs.csv")
# read
advs <- read_csv("advs.csv")
```
:::
:::

### Read and write Parquet

::: {.compare gutter="1.5rem"}
::: {.compare-col width="50%"}
**Polars**

```{python}
#| eval: false
advs.write_parquet("advs.parquet")
advs = pl.read_parquet("advs.parquet")
```
:::

::: {.compare-col width="50%"}
**tidyverse**

```{r}
#| eval: false
advs |> arrow::write_parquet("advs.parquet")
advs <- arrow::read_parquet("advs.parquet")
```
:::
:::

## Sources

- Polars Python API. <https://docs.pola.rs/api/python/stable/reference/>.
- Wickham H, Averick M, Bryan J, et al. Welcome to the tidyverse. *Journal of Open
  Source Software*, 2019;4(43):1686. <https://doi.org/10.21105/joss.01686>.
- Grolemund G, Wickham H. Dates and Times Made Easy with lubridate. *Journal of
  Statistical Software*, 2011;40(3):1-25. <https://doi.org/10.18637/jss.v040.i03>.
